---
title: "Qualitative Activity Recognition Prediction Model"
author: "Luis Luévano"
date: "24/05/2015"
output: html_document
---
## Synopsis
This project aims to build a model to predict how well a routine exercise is being performed by a study subject group. The subjects were asked to perform a weight lift exercise in 6 different ways which 1 of them is correctly done. To measure and track the movements, sensors were used in belt, arm, forearm and dumbbell. Detail data can be found in the [website](http://groupware.les.inf.puc-rio.br/har) where the dataset (that will be use here) comes from.

## Loading and preprocessing the data
The file containing the data can be found [here](https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv). Make sure to set your working directory where the pml-training.csv file is found: setwd("\<directory\>")
```{r}
rwDf <- read.csv("pml-training.csv", na.strings = c("NA", "#DIV/0!", "")) # raw data frame
```
Please note that "#DIV/0!" and "" (empty string) have been added as part of NA. These were found in an exploratory analysis using *summary* R base function and decided to add logic since the reading phase. 

### Exploratory Analysis
Using *summary* function (see below command), it is noticeable that we have three main data clean steps to do:
1. We have variables not needed for the study.
2. From the potential needed columns, some of them have a great number of NAs.
3. Due to the nature of the data (body movements measurements around axis) several columns must be correlated.
```{r, eval=FALSE}
summary(rwDf) # No results display, too many columns
```

### Cleaning data
Select only variables of interest that contains in the name **arm** and **dumbbell** plus the outcome to predict: **classe**.
```{r}
tomatch <- paste(c("arm", "dumbbell", "classe"), collapse = "|")
exDf <- rwDf[, grep(tomatch, colnames(rwDf), value = TRUE)] # excercise data frame
```
Lets see the different percentages of NAs in all dataset:
```{r}
unique(round(colMeans(is.na(exDf)), 2)) # get mean of NAs rounded by 2 decimals 
```
A lot of NAs in single columns, there is no value to imput values on them so is better to remove them:
```{r}
exDf <- exDf[, colMeans(is.na(exDf)) == 0] # only if no NAs found
```
Checking high correlation:
```{r}
M<-abs(cor(exDf[,1:39]))
diag(M) <- 0
which(M >= 0.9, arr.ind=T) 
```
Remove high correlated variables:
```{r}
exDf <- exDf[, !(names(exDf) %in% c("gyros_arm_y", "gyros_forearm_z", "gyros_dumbbell_x"))]
```
## Creating the predictive model
Lets create a train and test data frames using the final dataset of 36 predictors. A 60% train dataset seems reasonable having 19,622 records.
```{r, message=FALSE}
library(caret); library(randomForest)
inTrain = createDataPartition(y = exDf$classe, p = 0.6, list = FALSE)
training = exDf[inTrain,]
testing = exDf[-inTrain,]
```
Since our predictors are based on sensors, they are subject to noise input. A Random Forest algorithm will allow to decrease the variance caused by these. Lets use the default settings:
```{r, cache=TRUE}
modFit <- randomForest(classe~. , data = training)
```
Lets see how well it performs:
```{r}
confusionMatrix(testing$classe, predict(modFit,testing))
```

## Conclusions
The results from our very first model are quite good. The out of sample error is 0.0171, which is the error applied to the testing set given the prediction model. I believe this is enough for our purpose to predict how well a weight lift exercise is being done.

